---
title: "Spatial Analysis - Example"
author: "Dr Hazlienor"
date: "2024-10-25"
output: html_document
---

Example of analysis in spatial epidemiology using pre-existing dataset in R

#Geostatistical data

## Prepare environment 

### install packages
```{r}
install.packages(c("sf", "spData", "spdep", "ggplot2", "sp", "gstat"))
```

### Load the required libraries
```{r}
library(sf)         # Handling spatial data
library(gstat)      # Handling geostatistical data
library(spdep)      # Spatial dependency and autocorrelation
library(spData)
library(sp)         # Contains 'meuse' dataset
library(tmap)       # Data visualization
```

## Prepare data
We'll use the meuse dataset, which contains the spatial coordinates of soil samples with heavy metal concentrations. Imagine these metals affect public health (e.g., lead exposure affecting residents).
```{r}
# Load the dataset
data("meuse", package = "sp")

# Convert to spatial data frame
coordinates(meuse) <- ~x + y
proj4string(meuse) <- CRS("+init=epsg:28992")  # Dutch coordinate system

# Check the structure of the data
head(meuse)
```

## Mapping

# Create a simple map of zinc concentrations
```{r}
tm_shape(st_as_sf(meuse)) +
  tm_dots("zinc", title = "Zinc (ppm)", palette = "YlOrRd") +
  tm_layout(main.title = "Zinc Pollution in Study Area", main.title.size = 1.2)

```

## Variogram: Measure Spatial Dependence

A variogram helps analyze how zinc concentrations vary with distance, which is essential for geostatistical modeling.
```{r}
# Calculate experimental variogram
zinc_variogram <- variogram(zinc ~ 1, meuse)

# Plot the variogram
plot(zinc_variogram, main = "Experimental Variogram for Zinc")

```
We’ll fit a theoretical model (e.g., Spherical model) to the experimental variogram.
```{r}
# Fit a spherical variogram model
zinc_model <- fit.variogram(zinc_variogram, model = vgm(psill = 0.6, "Sph", range = 500, nugget = 0.2))

# Plot the fitted variogram model
plot(zinc_variogram, zinc_model, main = "Fitted Variogram Model (Zinc)")
```


Variogram analysis reveals how zinc concentrations vary with distance, which informs the spatial dependence of health risks.

## Kriging: Interpolating Pollution Levels Across the Region

Kriging is a geostatistical interpolation method used to predict values at unsampled locations.
```{r}
# Create a grid for interpolation
data("meuse.grid", package = "sp")
coordinates(meuse.grid) <- ~x + y
gridded(meuse.grid) <- TRUE


# Set the correct CRS for both objects
proj4string(meuse) <- CRS("+init=epsg:28992")
proj4string(meuse.grid) <- CRS("+init=epsg:28992")

# Perform ordinary kriging
zinc_kriging <- krige(zinc ~ 1, meuse, meuse.grid, model = zinc_model)

# Plot the kriging result
# Load the required library for raster manipulation
library(raster)

# Convert kriging results to a raster
zinc_raster <- raster(zinc_kriging, layer = "var1.pred")

# Plot the raster to confirm
plot(zinc_raster, main = "Kriged Zinc Levels")



```
Kriging interpolation predicts pollution levels in unsampled areas, highlighting potential hotspots.

# Areal data analysis

## Prepare Environment

### Install packages

Install necessary packages if not already installed
```{r}
install.packages(c("spDataLarge", "sf", "spdep", "tmap"))
```


### Load the libraries
```{r}
library(sf)           # Handling spatial data
library(spdep)        # Spatial dependency and autocorrelation
library(tmap)         # Thematic mapping
library(spDataLarge)  # Contains health-related datasets
```


## Prepare Data


The nc dataset from the sf package contains data on sudden infant death syndrome (SIDS) cases in North Carolina counties.
```{r}
# Load North Carolina dataset
nc <- st_read(system.file("shape/nc.shp", package = "sf"))


# Check the structure of the dataset
head(nc)
```

## Mapping

### Static Map

Plot the North Carolina dataset using tmap
```{r}
tm_shape(nc) +
  tm_polygons("AREA", title = "Area of Counties") +
  tm_layout(main.title = "North Carolina Counties", main.title.size = 1.5)

```

Plot SIDS rate
```{r}
tm_shape(nc) +
  tm_polygons("SID79", 
              title = "SIDS Rate (1979)", 
              palette = "Blues", 
              style = "quantile")

```
### Interactive map
```{r}
tmap_mode("view")  # Switch to interactive map

tm_shape(nc) +
  tm_polygons("SID79", 
              title = "SIDS Cases (1979)", 
              palette = "Reds")


```
## Spatial autocorrelation

### Global Moran's I

We create neighbors and weights matrices to calculate spatial autocorrelation using Moran's I.
```{r}
# Create spatial neighbors object
nc_neighbors <- poly2nb(nc)

# Convert neighbors to weights list
nc_weights <- nb2listw(nc_neighbors, style = "W")

# Calculate Moran's I for SIDS incidence in 1979
moran_result <- moran.test(nc$SID79, nc_weights)
print(moran_result)
```

Positive Moran's I suggests spatial clustering (neighboring counties with similar SIDS cases) at global scale.
A p-value < 0.05 indicates statistically significant clustering.

### Local Moran's I
```{r}
# Perform Local Moran's I analysis for SID79 incidence
local_moran <- localmoran(nc$SID79, nc_weights)

# Add the Local Moran's I values to the dataset
nc$local_I <- local_moran[, "Ii"]

# Classify regions based on the sign of Local Moran's I
nc$lisa_cluster <- ifelse(nc$local_I > 0, "Positive Autocorrelation", 
                          ifelse(nc$local_I < 0, "Negative Autocorrelation", 
                                 "No Autocorrelation"))

# View the first few rows to check the classifications
head(nc[, c("local_I", "lisa_cluster")])

```

Plot LISA map based on Local Moran's I value
```{r}
tm_shape(nc) +
  tm_polygons("lisa_cluster", 
              title = "LISA Map of SID79", 
              palette = c("red", "blue", "grey"), 
              labels = c("Positive Autocorrelation", "Negative Autocorrelation", "No Autocorrelation")) +
  tm_layout(main.title = "LISA Map of SID79 Incidence", main.title.size = 1.5)
```

Local Moran's I Values:

-Positive values: Indicates clustering of similar values (high-high or low-low).
-Negative values: Indicates spatial outliers (high-low or low-high).
-Zero/near-zero values: No clear autocorrelation.
-LISA Map: This map helps you visualize spatial patterns (positive or negative autocorrelation) without focusing on p-values.



## Spatial Regression
Fit a simple spatial autoregressive model to explore factors affecting SIDS cases.

Fit a spatial autoregressive model (SAR) using **spatialreg** package
SID79: SIDS cases in 1979
BIR79: Total births in 1979 (used as a predictor)
```{r}
library(spatialreg)
sar_model <- lagsarlm(SID79 ~ BIR79, data = nc, listw = nc_weights)
summary(sar_model)

```
Get predicted values
```{r}
# Get predicted values
nc$predicted_SID79 <- fitted(sar_model)

# Check the first few predicted values
head(nc$predicted_SID79)

```

Map the predicted values
```{r}
# Set up tmap mode to plot
tmap_mode("plot")

# Plot the predicted values using tmap
tm_shape(nc) +
  tm_polygons("predicted_SID79", title = "Predicted SID79", palette = "YlGnBu") +
  tm_layout(main.title = "Predicted SID79 Values from SAR Model", main.title.size = 1.5)

```

Calculate risk score
```{r}
average_predicted <- mean(nc$predicted_SID79, na.rm = TRUE)
nc$risk_score <- nc$predicted_SID79 / average_predicted
```

Map Relative Risk
```{r}
tm_shape(nc) +
  tm_polygons("risk_score", title = "Risk Score", palette = "RdYlBu") +
  tm_layout(main.title = "Risk Map Based on Predicted SID79", main.title.size = 1.5)
```
# Point Pattern Analysis

## Prepare Environment

### Install packages
```{r}
install.packages(c("sf", "spatstat", "spatstat.data", "ggplot2"))
```


### Load Library
```{r}
library(sf)              # Handling spatial data
library(spatstat)        # Point pattern analysis functions
library(spatstat.data)   # Contains the ohio dataset
library(ggplot2)         # Mapping
library(dplyr)           # Data manipulation
```
## Prepare data

The chorley object is a marked point pattern (class ppp), where each point corresponds to a cancer case. The marks indicate whether the patient was a lung cancer or larynx cancer patient
```{r}
# Load the dataset
data("chorley", package = "spatstat.data")

# Check the structure of the data
summary(chorley)
```


## Point Pattern Analysis

### Disease mapping

Plot the point pattern
```{r}
plot(chorley, main = "Cancer Cases in Chorley-Ribble Area")
```

### Kernel Density Estimation

Kernel density estimation helps visualize the intensity (hotspots) of cases across the study area.
```{r}
# Estimate the spatial intensity using a Gaussian kernel
density_chorley <- density(chorley, sigma = 1)

# Plot the kernel density estimate
plot(density_chorley, main = "Kernel Density of Cancer Cases")
```
This map shows the intensity of cancer cases, with brighter areas indicating higher concentrations.

### Quadrat Analysis

**Quadrat Test for Spatial Uniformity**

A quadrat test can determine if the distribution is spatially uniform or if some areas have significantly more cases than expected.

```{r}
# Perform a quadrat test (divide the area into 5x5 quadrats)
Q_test <- quadrat.test(chorley, nx = 5, ny = 5)

# Print the test result
print(Q_test)

# Plot the quadrat counts
plot(Q_test, main = "Quadrat Test for Chorley Cancer Cases")
```
If the p-value is significant, it indicates that the cases are not uniformly distributed and may be clustered in certain areas.


### K-function Analysis

The K-function is used to assess whether the points exhibit clustering, randomness, or regularity.
```{r}
# Compute the K-function for the point pattern
K_chorley <- Kest(chorley)

# Plot the K-function
plot(K_chorley, main = "K-function for Chorley Cancer Cases", legendargs = list(x = "topleft"))

```

If the observed K-function lies above the theoretical K-function (Poisson process), the pattern suggests clustering.

If it lies below, the pattern suggests dispersion.


```{r}

```

